\documentclass[sigconf]{acmart}

% -------------------------------------------------
% Metadata
% -------------------------------------------------
\title{Evaluating Durable Functions for Stateful Serverless Computing}

\author{Julio Sigüenza Pacheco}

\settopmatter{printacmref=false}
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}

% -------------------------------------------------
% Packages
% -------------------------------------------------
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}

\graphicspath{{figures/}}

% -------------------------------------------------
% Document
% -------------------------------------------------
\begin{document}

\begin{abstract}
Serverless platforms are traditionally stateless, forcing developers to manage
state explicitly through external storage systems. In December~2025, AWS
introduced Durable Functions for AWS Lambda, a new execution model that enables
stateful serverless workflows through checkpoint-and-replay semantics.

This paper presents a hands-on evaluation of AWS Lambda Durable Functions using
two representative workloads: a stateful counter and a video encoding pipeline.
We implement each workload in two variants: a durable execution and a
traditional Lambda-based baseline with explicit state storage. Using 20 runs for
the counter and 30 runs for the video pipeline, we measure execution time,
coordination overhead, and failure-handling behavior.

Our results show that durable execution completely eliminates explicit state
storage, reducing external coordination from an average of 5 operations per
counter execution and more than 60 operations per video pipeline run to zero,
while preserving strict sequential and actor-like semantics. Although durable
execution incurs higher end-to-end latency in the local simulator, it
significantly simplifies application logic and provides built-in fault
tolerance without manual retries or polling. These findings position AWS
Durable Functions as a practical realization of the actor model for serverless
computing.
\end{abstract}

\maketitle

% -------------------------------------------------
% Sections
% -------------------------------------------------
\input{sections/01-intro}
\input{sections/02-background}
\input{sections/03-arch}
\input{sections/04-evaluation}
\input{sections/05-actors}
\input{sections/06-conclusion}

% -------------------------------------------------
% References
% -------------------------------------------------
\bibliographystyle{ACM-Reference-Format}
\bibliography{refs}

\end{document}
